Approximation as Differentiation

### 2025-05-22 notes/stability.tex
Stability <-> left adjoint on slices

@@

\begin{definition}
  For a poset $X$ and element $x \in X$, the \emph{principal
    downset of $x$} is $\downset{x} = \{ x' \in X \mid x' \leq x \}$.
\end{definition}

@@

\begin{proof}
  If $f$ is stable, then define a left adjoint
  $g_x : \downset{f(x)} \to \downset{x}$ by setting $g_x(y)$ to be the
  minimal $x_0$ required by stability. (FIXME: this is monotone?). Let
  $x' \leq x$ and $y \leq f(x)$. Then if $g_x(y) \leq x'$, we have
  $y \leq f(g_x(y)) \leq f(x')$ by monotonicity of $f$ and the first
  part of stability. In the other direction, if we have
  $y \leq f(x')$, then by uniqueness we have $g_x(y) \leq x'$.

  TODO: left adjoint implies stability.
\end{proof}

>>>>

Stable functions arise in domain theory as an attempt to capture the
idea that any function describing a computational process must explore
its input in a way that allows identification of minimal parts of the
input that are needed for chosen approximations of the input. They
were invented by Berry \cite{berry86} in an attempt to capture
sequentiality, and were also instrumental in the discovery of Linear
Logic via Coherence Spaces \cite{girard}. A textbook description of
stable functions in the context of domain theory is given by Amadio
and Curien \cite{amadio-curien}. Here, we are not concerned with
completeness properties considered in domain theory. Stable functions
can be defined between any two partially ordered sets:

@@

\begin{definition}
  For a poset $X$ and $x \in X$, the \emph{principal
    downset of $x$} is $\downset{x} = \{ x' \in X \mid x' \leq x \}$.
\end{definition}

@@

\begin{proof}
  If $f$ is stable, then define a left adjoint
  $g_x : \downset{f(x)} \to \downset{x}$ by setting $g_x(y)$ to be the
  minimal $x_0$ required by stability. This is monotone: if
  $y \leq y'$, then we know that $y \leq y' \leq f(g_x(y'))$ by
  defintition of $g_x$, so $g_x(y) \leq g_x(y')$ by minimality of
  $g_x(y)$. For the adjointness, let $x' \leq x$ and $y \leq
  f(x)$. Then if $g_x(y) \leq x'$, we have
  $y \leq f(g_x(y)) \leq f(x')$ by monotonicity of $f$ and the first
  part of stability. In the other direction, if we have
  $y \leq f(x')$, then by uniqueness we have $g_x(y) \leq x'$.

  If, for every $x$, $f_x$ has a left adjoint $g_x$, then for any
  $x', y$ we have $y \leq f_x(x') \Leftrightarrow g_x(y) \leq x'$. So
  $g_x(y)$ is the element that satisfies $y \leq f(g_x(y))$, and it is
  minimal since if $y \leq f_x(x'_0)$ then $g_x(y) \leq x'_0$.
\end{proof}
